---
title: "Full list - COI"
author: "Emil Ellegaard Thomassen"
date: "`r Sys.Date()"
output: github_document
editor_options: 
  chunk_output_type: console
---


# Setup and loading

First, set up R and load packages

```{r Setup and loading packages}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 5.5,
	message = FALSE,
	warning = FALSE,
	autodep = TRUE,
	cache = TRUE,
	cache.lazy = FALSE,
	tidy = TRUE,
	tidy.opts = list(width.cutoff = 80)
)
rm(list=ls())
gc()
Dir.Base <- here::here()



# Installing packages as needed
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x, repos='http://cran.us.r-project.org', dependencies = TRUE)
  require(x, character.only = TRUE)
}
package_vec <- c("kableExtra", "tidyverse","png", "jpeg", "sp", "ncf","rgdal","MASS", "raster", "ggmap", "ggplot2", "BiodiversityR", "pheatmap", "knitr", "printr", "ade4", "vegan", "ROBITools", "ROBITaxonomy", "dplyr", "tidyr", "plotrix")
sapply(package_vec, install.load.package)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=FALSE, tidy.opts=list(width.cutoff=80),tidy=TRUE, fig.height=5.5)
options(digits=4, width = 90)

```

Then, metadata file is loaded

```{r Metadata}

samples.data = read.table("COI/MetaBarFlow/output/Metadata_MOLS2020_invertebrates_corrected_source.txt",
                        header=TRUE,row.names=1,na.strings="NA")
head(samples.data)

###The sample selector.

DungSamples = rep(FALSE,nrow(samples.data))
DungSamples = DungSamples | samples.data$type=="sample"
names(DungSamples)=rownames(samples.data)

##Galloway selector
GalSamples = rep(FALSE,nrow(samples.data))
GalSamples = GalSamples | samples.data$source=="galloway"
names(GalSamples)=rownames(samples.data)

##Exmoor selector
ExmSamples = rep(FALSE,nrow(samples.data))
ExmSamples = ExmSamples | samples.data$source=="exmoor"
names(ExmSamples)=rownames(samples.data)

##January selector
JanSamples = rep(FALSE,nrow(samples.data))
JanSamples = JanSamples | samples.data$month=="January"
names(JanSamples)=rownames(samples.data)

##February selector
FebSamples = rep(FALSE,nrow(samples.data))
FebSamples = FebSamples | samples.data$month=="February"
names(FebSamples)=rownames(samples.data)

##March selector
MarSamples = rep(FALSE,nrow(samples.data))
MarSamples = MarSamples | samples.data$month=="March"
names(MarSamples)=rownames(samples.data)

##April selector
AprSamples = rep(FALSE,nrow(samples.data))
AprSamples = AprSamples | samples.data$month=="April"
names(AprSamples)=rownames(samples.data)

##May selector
MaySamples = rep(FALSE,nrow(samples.data))
MaySamples = MaySamples | samples.data$month=="May"
names(MaySamples)=rownames(samples.data)

##June selector
JunSamples = rep(FALSE,nrow(samples.data))
JunSamples = JunSamples | samples.data$month=="June"
names(JunSamples)=rownames(samples.data)

##July selector
JulSamples = rep(FALSE,nrow(samples.data))
JulSamples = JulSamples | samples.data$month=="July"
names(JulSamples)=rownames(samples.data)

##August selector
AugSamples = rep(FALSE,nrow(samples.data))
AugSamples = AugSamples | samples.data$month=="August"
names(AugSamples)=rownames(samples.data)

##September selector
SepSamples = rep(FALSE,nrow(samples.data))
SepSamples = SepSamples | samples.data$month=="September"
names(SepSamples)=rownames(samples.data)

##October selector
OctSamples = rep(FALSE,nrow(samples.data))
OctSamples = OctSamples | samples.data$month=="October"
names(OctSamples)=rownames(samples.data)

##November selector
NovSamples = rep(FALSE,nrow(samples.data))
NovSamples = NovSamples | samples.data$month=="November"
names(NovSamples)=rownames(samples.data)

##December selector
DecSamples = rep(FALSE,nrow(samples.data))
DecSamples = DecSamples | samples.data$month=="December"
names(DecSamples)=rownames(samples.data)


```

Sample names are extracted:

```{r Extracting sample names}

#Read the filtered dataset
dataset <- import.metabarcoding.data('COI/MetaBarFlow/output/merged_table_all_23_11_22.txt')

pcr = factor(substr(rownames(dataset@samples),
                    nchar(rownames(dataset@samples))-1,
                    nchar(rownames(dataset@samples))))

samplenames=factor(substr(rownames(dataset@samples),
                          1,
                          nchar(rownames(dataset@samples))-2))

##Preparing selector vectors

#DungSamples
samples.DungSamples = DungSamples[as.character(samplenames)]
samples.DungSamples[is.na(samples.DungSamples)]=FALSE
names(samples.DungSamples)=rownames(dataset@samples)
samples.DungSamples

#Galloway Samples
samples.GalSamples = GalSamples[as.character(samplenames)]
samples.GalSamples[is.na(samples.GalSamples)]=FALSE
names(samples.GalSamples)=rownames(dataset@samples)

#Exmoor Samples
samples.ExmSamples = ExmSamples[as.character(samplenames)]
samples.ExmSamples[is.na(samples.ExmSamples)]=FALSE
names(samples.ExmSamples)=rownames(dataset@samples)

#January Samples
samples.JanSamples = JanSamples[as.character(samplenames)]
samples.JanSamples[is.na(samples.JanSamples)]=FALSE
names(samples.JanSamples)=rownames(dataset@samples)

#February Samples
samples.FebSamples = FebSamples[as.character(samplenames)]
samples.FebSamples[is.na(samples.FebSamples)]=FALSE
names(samples.FebSamples)=rownames(dataset@samples)

#March Samples
samples.MarSamples = MarSamples[as.character(samplenames)]
samples.MarSamples[is.na(samples.MarSamples)]=FALSE
names(samples.MarSamples)=rownames(dataset@samples)

#April Samples
samples.AprSamples = AprSamples[as.character(samplenames)]
samples.AprSamples[is.na(samples.AprSamples)]=FALSE
names(samples.AprSamples)=rownames(dataset@samples)

#May Samples
samples.MaySamples = MaySamples[as.character(samplenames)]
samples.MaySamples[is.na(samples.MaySamples)]=FALSE
names(samples.MaySamples)=rownames(dataset@samples)

#June Samples
samples.JunSamples = JunSamples[as.character(samplenames)]
samples.JunSamples[is.na(samples.JunSamples)]=FALSE
names(samples.JunSamples)=rownames(dataset@samples)

#July Samples
samples.JulSamples = JulSamples[as.character(samplenames)]
samples.JulSamples[is.na(samples.JulSamples)]=FALSE
names(samples.JulSamples)=rownames(dataset@samples)

#August Samples
samples.AugSamples = AugSamples[as.character(samplenames)]
samples.AugSamples[is.na(samples.AugSamples)]=FALSE
names(samples.AugSamples)=rownames(dataset@samples)

#September Samples
samples.SepSamples = SepSamples[as.character(samplenames)]
samples.SepSamples[is.na(samples.SepSamples)]=FALSE
names(samples.SepSamples)=rownames(dataset@samples)

#October Samples
samples.OctSamples = OctSamples[as.character(samplenames)]
samples.OctSamples[is.na(samples.OctSamples)]=FALSE
names(samples.OctSamples)=rownames(dataset@samples)

#November Samples
samples.NovSamples = NovSamples[as.character(samplenames)]
samples.NovSamples[is.na(samples.NovSamples)]=FALSE
names(samples.NovSamples)=rownames(dataset@samples)

#December Samples
samples.DecSamples = DecSamples[as.character(samplenames)]
samples.DecSamples[is.na(samples.DecSamples)]=FALSE
names(samples.DecSamples)=rownames(dataset@samples)



###The extraction control selector. Their names start with "CNE"
ext_con = rep(FALSE,
               nrow(dataset@samples))

ext_con[grep('^CNE',
              rownames(dataset@samples))]=TRUE

###The PCR control selector. Their names start with "NTC"
ntc = rep(FALSE,
               nrow(dataset@samples))

ntc[grep('^NTC',
              rownames(dataset@samples))]=TRUE

###The field blank control selector. Their names end with "K"
FB = rep(FALSE,
               nrow(dataset@samples))

FB[grep('K',
              rownames(dataset@samples))]=TRUE

##Merging the metadata database with the metabarcoding object

#Source (Exmoor/galloway)
source=as.character(samples.data[as.character(samplenames),]$source)
source=factor(source)

#Months
month=as.character(samples.data[as.character(samplenames),]$month) 
month=factor(month)

#Season
season=as.character(samples.data[as.character(samplenames),]$season) 
season=factor(season)

#Habitat
habitat=as.character(samples.data[as.character(samplenames),]$habitat) 
habitat=factor(habitat)

#Latitude
lat=samples.data[as.character(samplenames),]$lat 

#Longitude
lon=samples.data[as.character(samplenames),]$lon 


###All the information are collected in a data.frame associated to the data ###element of the metabarcoding object.

m=dataset

#For Dung samples
m@samples = cbind(dataset@samples,
                       samples.DungSamples,
                       samples.GalSamples,
                       samples.ExmSamples,
                       samples.JanSamples,
                       samples.FebSamples,
                       samples.MarSamples,
                       samples.AprSamples,
                       samples.MaySamples,
                       samples.JunSamples,
                       samples.JulSamples,
                       samples.AugSamples,
                       samples.SepSamples,
                       samples.OctSamples,
                       samples.NovSamples,
                       samples.DecSamples,
                       ext_con,
                       ntc,
                       FB,
                       source,
                       month,
                       season,
                       habitat,
                       lat,
                       lon,
                       name=samplenames,
                       repeats=pcr
                       )
dataset=m


```



```{r Read count per sample}

par(mfrow=c(1,2))

hist(rowSums(dataset@reads[dataset@samples$samples.DungSamples,]),
     breaks=50,xlab="Reads",main="Read counts")

hist(log10(rowSums(dataset@reads[dataset@samples$samples.DungSamples,])),
     breaks=50,xlab="Log(reads)",main="Log of read counts")



options(digits=10)
tot_reads<-sum(dataset@reads)
reads_FB<-sum(dataset@reads[dataset@samples$FB,])
reads_CNE<-sum(dataset@reads[dataset@samples$ext_con,])
reads_NTC<-sum(dataset@reads[dataset@samples$ntc,])
tot_reads_samples<-sum(dataset@reads[dataset@samples$samples.DungSamples,])

slices<-c(tot_reads_samples,reads_FB,reads_CNE,reads_NTC)
lbs<-c("Samples", "Field Blanks", "Extraction Blanks", "PCR Blanks")
pct <- slices/sum(slices)*100

par(mfrow=c(1,1))
pie(slices,labels=lbs, main="Read distribution", col = c("darkseagreen","darkgoldenrod","blue", "red"))

#Save these values in a table

reads_dist<-data.frame("tot_reads"=tot_reads,"reads_FB"=reads_FB,"reads_CNE"=reads_CNE,"reads_NTC"=reads_NTC,"tot_reads_samples"=tot_reads_samples)



```

```{r ASV count per sample}



par(mfrow=c(1,2))

presence = dataset@reads[dataset@samples$samples.DungSamples,] > 0

hist(rowSums(presence),
     breaks=50,xlab="ASVs",main="ASVs")
hist(log10(rowSums(presence)),                                      
     breaks=50,xlab="Log(ASVs)",main="ASVs")

```

```{r Scatterplot reads vs motus}

par(mfrow=c(1,1))

  presence = dataset@reads[dataset@samples$samples.DungSamples,] > 0       
  plot(rowSums(dataset@reads[dataset@samples$samples.DungSamples,]),       
     rowSums(presence),
     log="xy",xlab="Reads",ylab="ASVs",main="ASVs vs reads")

```

```{r Check ntc contaminants}

par(mfrow=c(2,4))

  negControl <- dataset@reads[dataset@samples$ntc,]
  if(is.null(dim(negControl))) negControl <- matrix(negControl, nrow=1) #vector to matrix
  cat("Negative control for ","MOLS2020_invertebrates",": NTC\n")
  print(dataset@motus[colSums(negControl)>0,  c('count','final.otu')])

```

There are ASVs present in controls, thus these are removed from samples if they are present in higher number in any control compared to any sample:


```{r Remove ntc contamination}

par(mfrow=c(1,1))

contam=dataset@motus[colSums(dataset@reads[dataset@samples$ntc,])>0,
                      c('count','final.otu')]

if(length(rownames(contam))>1){
  contam=cbind(contam,
             max.sample=apply(dataset@reads[dataset@samples$samples.DungSamples,  #CHANGE Stream_water
                                             rownames(contam)],2,max),
             max.ntc=apply(dataset@reads[dataset@samples$ntc,
                                               rownames(contam)],2,max))}
if(length(rownames(contam))==1){
 X1 = dataset@reads[dataset@samples$samples.DungSamples,rownames(contam)]         #CHANGE Stream_water
 X1 <- matrix(X1,ncol=1)
 X2 = dataset@reads[dataset@samples$ntc,rownames(contam)]
 X2 <- matrix(X2,ncol=1)  
 contam=cbind(contam,
             max.sample=apply(X1,2,max),
             max.ntc=apply(X2,2,max))}

knitr::kable(contam[order(-contam$max.ntc),])

##Remove all MOTUs that are more frequent in a pcr control than in a sample
contaminant.motus = rownames(contam)[contam$max.ntc>=contam$max.sample]
contaminant.motus

write.table(contam, file='COI/Local_analysis/output/textfiles/MOLS2020_invertebrates_contam_ntc_total.txt', quote=FALSE, sep='\t', col.names = NA,row.names=TRUE)



m = dataset[,!(colnames(dataset@reads) %in% rownames(contam)[contam$max.ntc>=contam$max.sample])]
m@motus$count=colSums(m@reads)

m=m[rowSums(m@reads)>0,]

dataset=m

presence = m@reads[m@samples$samples.DungSamples,] > 0                                       
plot(rowSums(m@reads[m@samples$samples.DungSamples,]),                                       
     rowSums(presence),
     log="xy",xlab="Reads",ylab="ASVs",main="ASVs vs reads after NTC filter")

```

```{r Check Field controls contaminants}

par(mfrow=c(2,4))

  fieldBlank <- dataset@reads[dataset@samples$FB,]
  if(is.null(dim(fieldBlank))) fieldBlank <- matrix(fieldBlank, nrow=1) #vector to matrix
  cat("Negative control for ","MOLS2020_invertebrates",": FB\n")
  print(dataset@motus[colSums(fieldBlank)>0,  c('count','final.otu')])

 
  
```

There are ASVs present in controls, thus these are removed from samples if they are present in higher number in any control compared to any sample:


```{r Remove FB contamination}

par(mfrow=c(1,1))

contam=dataset@motus[colSums(dataset@reads[dataset@samples$FB,])>0,
                      c('count','final.otu')]

if(length(rownames(contam))>1){
  contam=cbind(contam,
             max.sample=apply(dataset@reads[dataset@samples$samples.DungSamples,         #CHANGE Stream_water
                                             rownames(contam)],2,max),
             max.FB=apply(dataset@reads[dataset@samples$FB,
                                               rownames(contam)],2,max))}
if(length(rownames(contam))==1){
 X1 = dataset@reads[dataset@samples$samples.DungSamples,rownames(contam)]                #CHANGE Stream_water
 X1 <- matrix(X1,ncol=1)
 X2 = dataset@reads[dataset@samples$FB,rownames(contam)]
 X2 <- matrix(X2,ncol=1)  
 contam=cbind(contam,
             max.sample=apply(X1,2,max),
             max.FB=apply(X2,2,max))}

knitr::kable(contam[order(-contam$max.FB),])

##Remove all MOTUs that are more frequent in a Field control than in a sample
contaminant.motus = rownames(contam)[contam$max.FB>=contam$max.sample]
contaminant.motus

write.table(contam, file='COI/Local_analysis/output/textfiles/MOLS2020_invertebrates_contam_FB_total.txt', quote=FALSE, sep='\t', col.names = NA,row.names=TRUE)

m = dataset[,!(colnames(dataset@reads) %in% rownames(contam)[contam$max.FB>=contam$max.sample])]
m@motus$count=colSums(m@reads)

m=m[rowSums(m@reads)>0,]

dataset=m

presence = m@reads[m$samples$samples.DungSamples,] > 0                                           
plot(rowSums(m@reads[m$samples$samples.DungSamples,]),                                           
     rowSums(presence),
     log="xy",xlab="Reads",ylab="ASVs",main="ASVs vs reads after FB filter")

```

```{r Check cne contaminants}

  extControl <- dataset@reads[dataset@samples$ext_con,]
  if(is.null(dim(extControl))) extControl <- matrix(extControl, nrow=1) #vector to matrix
  cat("Extraction control for ","MOLS2020_invertebrates",": CNE\n")
  print(dataset@motus[colSums(extControl)>0,  c('count','final.otu')])

```

There are ASVs present in controls, thus these are removed from samples if they are present in higher number in any control compared to any sample:


```{r Remove cne contaminants}

par(mfrow=c(1,2))

contam=dataset@motus[colSums(dataset@reads[dataset@samples$ext_con,])>0,
                      c('count','final.otu')]



if(length(rownames(contam))>1){
  contam=cbind(contam,
             max.sample=apply(dataset@reads[dataset@samples$samples.DungSamples,              #CHANGE Stream_water
                                             rownames(contam)],2,max),
             max.ext_con=apply(dataset@reads[dataset@samples$ext_con,
                                               rownames(contam)],2,max))}
if(length(rownames(contam))==1){
 X1 = dataset@reads[dataset@samples$samples.DungSamples,rownames(contam)]                     #CHANGE Stream_water
 X1 <- matrix(X1,ncol=1)
 X2 = dataset@reads[dataset@samples$ext_con,rownames(contam)]
 X2 <- matrix(X2,ncol=1)  
 contam=cbind(contam,
             max.sample=apply(X1,2,max),
             max.ext_con=apply(X2,2,max))}

knitr::kable(contam[order(-contam$max.ext_con),])

#Remove all MOTUs that are more frequent in an extraction control than in a sample
contaminant.motus = rownames(contam)[contam$max.ext_con>=contam$max.sample]
contaminant.motus

write.table(contam, file='COI/Local_analysis/output/textfiles/MOLS2020_invertebrates_contam_ext_total.txt', quote=FALSE, sep='\t', col.names = NA,row.names=TRUE)

m = dataset[,!(colnames(dataset@reads) %in% rownames(contam)[contam$max.ext_con>=contam$max.sample])]
m@motus$count=colSums(m@reads)
m=m[rowSums(m@reads)>0,]

repeats =  table(dataset@samples[dataset@samples$samples.DungSamples,]$name)[unique(dataset@samples[dataset@samples$samples.DungSamples,'name'])]                                                                                 
dataset=m


```

```{r Remove singletons}

#Removes singletons



  m=dataset
  reads = m@reads
  for (i in unique(m@samples$name)){
    idxName = grepl(unlist(i), row.names(m@samples))
    if(sum(idxName)>1)
      checkValues = colSums(reads[idxName,]>0)
      # print(unique(dataset@motus$scientific_name[names(checkValues)])) # not working?
    if(sum(idxName)==1)
      checkValues = (reads[idxName,]>0)
      idxEqualsOne = which(checkValues == 1)
      reads[idxName, idxEqualsOne] = 0
  }
  m@reads <- reads
  m@motus$count=colSums(m@reads)
  m=m[rowSums(m@reads)>0,]
  dataset = m  
  
  
  
    

```


```{r Revove sequences appearing in only one PCR replicate}


m=dataset
reads = m@reads



for (name in unique(m@samples$name)) {
    idxVec <-  which(m@samples$name == unlist(name))
    for (i in 1:ncol(reads)) {
      n<-0
      for (idx in idxVec) {
        if (reads[idx, i]!=0) {
          n<-n+1
        }
      }
    if (n==1) {
      for (idx in idxVec) {
        reads[idx,i]=0
    }  
    }
}
}

m@reads <- reads
m@motus$count=colSums(m@reads)
m=m[rowSums(m@reads)>0,]
dataset = m



```

```{r Aggregate PCR replicates}

MOLS2020_COI_final = ROBITools:::aggregate.metabarcoding.data(dataset,
by=list(sample=dataset@samples$name),
FUN=sum)
MOLS2020_COI_final = MOLS2020_COI_final[,colSums(MOLS2020_COI_final@reads)>0]
MOLS2020_COI_final@motus$count=colSums(MOLS2020_COI_final@reads)
MOLS2020_COI_final=MOLS2020_COI_final[rowSums(MOLS2020_COI_final@reads)>0,]
MOLS2020_COI_final=MOLS2020_COI_final[MOLS2020_COI_final@samples$samples.DungSamples]
MOLS2020_COI_final$sqrt = sqrt(MOLS2020_COI_final@reads)

no.rare=MOLS2020_COI_final

#Read counts before rarefy
options(digits=10)
tot_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.DungSamples,])
gal_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.GalSamples,])
exm_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.ExmSamples,])

jan_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.JanSamples,])
feb_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.FebSamples,])
mar_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.MarSamples,])
apr_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.AprSamples,])
may_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.MaySamples,])
jun_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.JunSamples,])
jul_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.JulSamples,])
aug_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.AugSamples,])
sep_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.SepSamples,])
oct_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.OctSamples,])
nov_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.NovSamples,])
dec_reads_no_rare<-sum(no.rare@reads[no.rare@samples$samples.DecSamples,])


dataset <- no.rare

```

```{r Generating input df and create genera list with reads}

#Generate df with ID and reads


reads<-as.data.frame(dataset@reads)

reads<-t(reads)
otu<-as.character(dataset@motus$final.otu)
kingdom<-as.character(dataset@motus$kingdom_name)
phylum<-as.character(dataset@motus$phylum_name)
class<-as.character(dataset@motus$class_name)
order<-as.character(dataset@motus$order_name)
family<-as.character(dataset@motus$family_name)
genus<-as.character(dataset@motus$genus_name)
species<-as.character(dataset@motus$species_name)
dung<-as.character(dataset@motus$dung.associated)
type<-as.character(dataset@motus$type)
row.names(reads)<-otu

total<-rowSums(reads)

df<-as.data.frame(cbind(reads,total))
df<-as.data.frame(cbind(df,kingdom,phylum,class,order,family,genus))
df<-df[order(df$total, decreasing=TRUE),]


#Export list
write.csv2(df,"COI/Local_analysis/output/tables/Full_list_metazoa.csv", row.names=TRUE)


```
