---
title: "GLLVM with richness data, COI"
author: "Emil Ellegaard Thomassen"
date: "`r Sys.Date()"
output: html_document
editor_options: 
  chunk_output_type: console
---

The following script performs Generalized linear latent variable models of species richness for each order (gllvm).

The package gllvm is used. (https://jenniniku.github.io/gllvm/index.html) (Niku *et al.*, 2019)

# Data:

First, all dung-associated species are selected, as other species would represent random visits in dung pads, or contamination through airborne DNA or carry-over by other organisms.

Abund: Abundance matrix. Matrix with samples as rows and orders as columns. "Abundance values" are species counts (including ASVs identified at higher taxonomic levels, e.g., "Aphodius sp.12 (sphacelatus/prodromus))

Envs: Environmental variables. Data frame with samples as rows, and the following environmental variables as columns:
- Source (values: Exmoor/Galloway)
- Season (values: Early/Mid/Late)
- Habitat (values: Forest/Open)



# Setup

First, R environment is set up, and packages are loaded:

```{r Setup and load packages, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 5.5,
	message = FALSE,
	warning = FALSE,
	autodep = TRUE,
	cache = TRUE,
	cache.lazy = FALSE,
	tidy = TRUE,
	tidy.opts = list(width.cutoff = 80)
)
rm(list=ls())
gc()
setwd(here::here())



# Installing packages as needed
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x, repos='http://cran.us.r-project.org', dependencies = TRUE)
  require(x, character.only = TRUE)
}
package_vec <- c("kableExtra", "tidyverse", "ggplot2", "knitr", "printr", "vegan", "ROBITools", "ROBITaxonomy", "dplyr", "tidyr","gllvm","ggalt","ggpubr","formatR")
sapply(package_vec, install.load.package)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=FALSE, tidy.opts=list(width.cutoff=80),tidy=TRUE, fig.height=5.5)
options(digits=4, width = 90)

```

# Load data

Data is loaded and re-organized to match input to the gllvm() function:

```{r Load data, echo=TRUE}

#Load dataset for extracting environmental variables
dataset<-readRDS("COI/Local_analysis/output/RDS_files/dataset_final_all_dung")

#dataset<-readRDS("../output/RDS_files/dataset_final_all_dung") #only when knitting


#Load richness data
abund<-readRDS("COI/Local_analysis/output/RDS_files/richness_each_order")

#abund<-readRDS("../output/RDS_files/richness_each_order") #only when knitting


#Extract environmental variables
envs<-data.frame("source"=dataset@samples$source, "season"=dataset@samples$season, "habitat"=dataset@samples$habitat, "month"=dataset@samples$month)

#Change season-column to factor with levels "Early", "Mid" & "Late" instead of spring, summer, autumn and winter

envs<-envs %>%
  dplyr::mutate(season=factor(season, levels=c("Early","Mid","Late")))

envs$season[which(envs$month=="January")]<-"Early"
envs$season[which(envs$month=="February")]<-"Early"
envs$season[which(envs$month=="March")]<-"Early"
envs$season[which(envs$month=="April")]<-"Early"
envs$season[which(envs$month=="May")]<-"Mid"
envs$season[which(envs$month=="June")]<-"Mid"
envs$season[which(envs$month=="July")]<-"Mid"
envs$season[which(envs$month=="August")]<-"Mid"
envs$season[which(envs$month=="September")]<-"Late"
envs$season[which(envs$month=="October")]<-"Late"
envs$season[which(envs$month=="November")]<-"Late"
envs$season[which(envs$month=="December")]<-"Late"

#Change class of abundance data to integers
abund<-as.data.frame(abund)
abund<-lapply(abund, as.integer)
abund<-as.data.frame(abund)

#Set rownames of both data frames to sample names
rownames(envs)<-dataset@samples$name
rownames(abund)<-dataset@samples$name

#Create list of abundance data and environmental variables
df_list<-list("abund"=abund,"env"=envs)

#Drop the month column in environmental variables, as it was only used to change season values
df_list$env<-df_list$env[,c(1:3)]

#Drop "control" as factor level for source column (remnant from the initial filtering steps)
df_list$env$source<-droplevels(df_list$env$source)


#Define inputs (y=abundance matrix, X=data frame of environmental variables)
y <- as.matrix(df_list$abund)
X <- df_list$env


```

# Modeling - GLLVM

First, modeling is performed with different distributions

## Gaussian distribution

```{r GLLVM - gaussian, echo=TRUE}

# Gaussian
fitG <- gllvm(y = y, X = X, formula = y ~ source + season + habitat, family = "gaussian")

par(mfrow = c(2,2))
plot(fitG)
AIC.fitG<-AIC(fitG)
AIC.fitG


```

AIC value is `r AIC.fitG` - diagnostics plots does not look good! 

## Poisson distribution

```{r GLLVM - poisson, echo=TRUE}
# Poisson distribution
fitP <- gllvm(y = y, X = X, formula = y ~ source + season + habitat, family = poisson())

par(mfrow = c(2,2))
plot(fitP)
AIC.fitP<-AIC(fitP)
AIC.fitP

```

AIC value is `r AIC.fitP` - diagnostics plots looks better, but seems to be an overdispersion problem! 

## Negative binomial distibution

```{r GLLVM - negative binomial, echo=TRUE}

# Negative binomial
fitNB <- gllvm(y = y, X = X, formula = y ~ source + season + habitat, family = "negative.binomial")

par(mfrow = c(2,2))
plot(fitNB)
AIC.fitNB<-AIC(fitNB)
AIC.fitNB


```

AIC value is `r AIC.fitNB` - diagnostics plots shows that NB model seems to have taken care of the overdispersion problem! 

# Test of number of latent variables

Now several models with increasing number of latent variables are tested to find the optimum

```{r Test of number of LVs, echo=TRUE}

AICs.lv.test<-NULL
LVs<-NULL
for (i in c(1:5)) {
  mod<-gllvm(y = y, X = X, formula = y ~ source + season + habitat, family = "negative.binomial", num.lv = i)
  AICs.lv.test<-append(AICs.lv.test, AIC(mod))
  LVs<-append(LVs, i)
}

par(mfrow=c(1,1))
plot(LVs,AICs.lv.test, main="AIC with increasing Latent variables", xlab="Number of LVs", ylab="AIC", col="red", pch=19)

```

It seems like AIC increases almost linearly with increasing latent variables. I am not sure how to make sense of this, as i would have expected AIC to decrease with increasing number of latent variables? But seems like 1 would be the best model from AIC.

# Including interactions between environmental variables

Try to include interactions between predictors in the model

```{r GLLVM - NB and interactions, echo=TRUE}

# Including interaction term
fitNB.int<-gllvm(y = y, X = X, formula = y ~ source * season * habitat, family = "negative.binomial", num.lv = 1)

par(mfrow = c(2,2))
plot(fitNB.int)
AIC.fitNB.int<-AIC(fitNB.int)

#Compare to model without interactions
fitNB.noint<-gllvm(y = y, X = X, formula = y ~ source + season + habitat, family = "negative.binomial", num.lv = 1)

AIC.fitNB.noint<-AIC(fitNB.noint)


```

AIC without interaction is: `r AIC.fitNB.noint`
AIC with interaction is: `r AIC.fitNB.int`

Higher AIC indicate that including interactions does not improve the model.

# Test if some orders migth be outliers

If some orders are only present in very few samples, they might disproportionately drive the analysis. Try to investigate this, and remove outlier orders.

```{r Identify outlier orders, echo=TRUE}

#Calculate number of samples with presence of each order
nsamples<-NULL
for (i in 1:ncol(y)) {nsamples<-append(nsamples,length(y[which(y[,i]>0),i]))}

out<-cbind(colnames(y),nsamples,colSums(y))
out

```

There seems to be several orders that are only present in very few samples. Orders present in fewer than 20 samples are removed

```{r Remove outlier orders, echo=TRUE}

y.subs<-y[,which(nsamples>20)]

```

Now, modeling is performed with the dataset without the outlier orders

# Modeling - without outliers

## Poisson

```{r GLLVM - without outliers - poisson, echo=TRUE}

# Poisson distribution
sfitP <- gllvm(y = y.subs, X = X, formula = y ~ source + season + habitat, family = poisson())

par(mfrow = c(2,2))
plot(sfitP)
AIC.sfitP<-AIC(sfitP)
AIC.sfitP


```

Now a poisson distribution looks better than before - seems that the zero-inflation might be caused by some of the underrepresented orders, which are now removed as outliers. 

AIC is `r AIC.sfitP`

Still maybe a little overdispersion, so negative binomial is tried.

## Negative binomial

```{r GLLVM - without outliers - negative binomial, echo=TRUE}

# Negative binomial
sfitNB <- gllvm(y = y.subs, X = X, formula = y ~ source + season + habitat, family = "negative.binomial")

par(mfrow = c(2,2))
plot(sfitNB)
AIC.sfitNB<-AIC(sfitNB)
AIC.sfitNB

```

Looks generally good from diagnostics plots, but AIC (`r AIC.sfitNB`) is higher than the poisson model. Thus the poisson model is chosen.

# Test number of latent variables

Models with increasing number of LVs are performed, and AICs are plotted.

```{r Test number of lvs - without outliers, echo=TRUE}
AICs.lv.test<-NULL
LVs<-NULL
for (i in c(1:5)) {
  mod<-gllvm(y = y.subs, X = X, formula = y ~ source + season + habitat, family = poisson(), num.lv = i)
  AICs.lv.test<-append(AICs.lv.test, AIC(mod))
  LVs<-append(LVs, i)
}

par(mfrow=c(1,1))
plot(LVs,AICs.lv.test, main="AIC with increasing Latent variables (without outliers)", xlab="Number of LVs", ylab="AIC", col="red", pch=19)
```

Including 2 latent variables seems to be the best model. I have tried this several times, and the AIC values are not consistent between model fits. Sometimes 2 is best, sometimes 3.

I chose 2 for now, but we can decide to change if you think so. The results are pretty much the same.

# Including interactions

Test if including interactions between predictors improve model performance

```{r Test - including interactions - without outliers, echo=TRUE}
# Including interaction term
sfitP.int<-gllvm(y = y.subs, X = X, formula = y ~ source * season * habitat, family = "negative.binomial", num.lv = 2)

par(mfrow = c(2,2))
plot(sfitP.int)
AIC.sfitP.int<-AIC(sfitP.int)

sfitP.noint<-gllvm(y = y.subs, X = X, formula = y ~ source + season + habitat, family = "negative.binomial", num.lv = 2)
AIC.sfitP.noint<-AIC(sfitP.noint)

```

AIC without interaction: `r AIC.sfitP.noint`
AIC with interaction: `r AIC.sfitP.int`

Inclusion of interactions seem to improve model performance.

# Inclusion of random row effect

```{r GLLVM - random row effect, echo=TRUE}

#Random row effect included
sfitP.lv2.RX<-gllvm(y = y.subs, X = X, formula = y ~ source * season * habitat, family = "negative.binomial", num.lv = 2, row.eff = "random")


par(mfrow = c(2,2))
plot(sfitP.lv2.RX)                    
AIC.sfitP.lv2.RX<-AIC(sfitP.lv2.RX)
AIC.sfitP.lv2.RX

```

Inclusion of random row effect does not increase model performance:

AIC with row-effects: `r AIC.sfitP.lv2.RX`
AIC without: `r AIC.sfitP.noint`

Maybe we should include anyway, when our samples are not truly independent?



```{r Test simpler models, echo=TRUE}

#Random row effect included
sfitP.so_se<-gllvm(y = y.subs, X = X, formula = y ~ source + season, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.so_se)
sfitP.so_ha<-gllvm(y = y.subs, X = X, formula = y ~ source + habitat, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.so_ha)
sfitP.se_ha<-gllvm(y = y.subs, X = X, formula = y ~ season + habitat, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.se_ha)
sfitP.se<-gllvm(y = y.subs, X = X, formula = y ~ season, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.se)
sfitP.so<-gllvm(y = y.subs, X = X, formula = y ~ source, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.so)
sfitP.ha<-gllvm(y = y.subs, X = X, formula = y ~ habitat, family = "negative.binomial", num.lv = 2, row.eff = "random")
AIC(sfitP.ha)


```


None of these simpler model better than the one with all 3 predictors.

So best model is a model with poisson distribution, without outlier orders, including 2 latent variables, source, season and habitat as predictors, and including a random row-effect, and no interactions between predictors

# Best model

```{r Best model, echo=TRUE}

#Best model

fitBEST<-gllvm(y = y.subs, X = X, formula = y ~ source + season + habitat, family = poisson(), num.lv = 2, row.eff = "random")

fitBEST2<-gllvm(y = y.subs, X = X, formula = y ~ source * season * habitat, family = poisson(), num.lv = 2, row.eff = "random")

sumBEST<-summary(fitBEST)
sumBEST

```


# Create coefficient plots

First input is created from the summary of the model, and by extracting confidence intervals

```{r Create input for plotting, echo=TRUE}

#Gather all parameters and confidence intervals in df

conf<-as.data.frame(confint(fitBEST))



df.params<-as.data.frame(cbind(fitBEST$params$beta0,fitBEST$params$Xcoef,
                               conf$`2.5 %`[30:38],
                               conf$`97.5 %`[30:38],
                               conf$`2.5 %`[39:47],
                               conf$`97.5 %`[39:47],
                               conf$`2.5 %`[48:56],
                               conf$`97.5 %`[48:56],
                               conf$`2.5 %`[57:65],
                               conf$`97.5 %`[57:65]
))
                               
                               
colnames(df.params)<-c("intercept", "coef.source.GAL","coef.season.MID","coef.season.LATE", "coef.habitat.OPEN",
                          "conf.low.source.GAL","conf.high.source.GAL","conf.low.season.MID","conf.high.season.MID","conf.low.season.LATE","conf.high.season.LATE", "conf.low.habitat.OPEN", "conf.high.habitat.OPEN")

#Check if zero is included in 95% confint
signif.GAL<-NULL
for(i in 1:nrow(df.params)) {if (between(0, df.params$conf.low.source.GAL[i], df.params$conf.high.source.GAL[i])) {signif.GAL<-append(signif.GAL,"no diff.")}
                             else {if (df.params$conf.high.source.GAL[i]<0) {signif.GAL<-append(signif.GAL,"higher richness horse")}
                               else {signif.GAL<-append(signif.GAL,"higher richness cattle")}
                               }
}

signif.MID<-NULL
for(i in 1:nrow(df.params)) {if (between(0, df.params$conf.low.season.MID[i], df.params$conf.high.season.MID[i])) {signif.MID<-append(signif.MID,"no diff.")}
                             else {if (df.params$conf.high.season.MID[i]<0) {signif.MID<-append(signif.MID,"higher richness early")}
                               else {signif.MID<-append(signif.MID,"higher richness mid")}
                               }
}

signif.LATE<-NULL
for(i in 1:nrow(df.params)) {if (between(0, df.params$conf.low.season.LATE[i], df.params$conf.high.season.LATE[i])) {signif.LATE<-append(signif.LATE,"no diff.")}
                             else {if (df.params$conf.high.season.LATE[i]<0) {signif.LATE<-append(signif.LATE,"higher richness early")}
                               else {signif.LATE<-append(signif.LATE,"higher richness late")}
                               }
}

signif.OPEN<-NULL
for(i in 1:nrow(df.params)) {if (between(0, df.params$conf.low.habitat.OPEN[i], df.params$conf.high.habitat.OPEN[i])) {signif.OPEN<-append(signif.OPEN,"no diff.")}
                             else {if (df.params$conf.high.habitat.OPEN[i]<0) {signif.OPEN<-append(signif.OPEN,"higher richness forest")}
                               else {signif.OPEN<-append(signif.OPEN,"higher richness open")}
                               }
}


df.params<-cbind(df.params, "signif.GAL"=as.factor(signif.GAL),"signif.season.MID"=as.factor(signif.MID),"signif.season.LATE"=as.factor(signif.LATE),"signif.OPEN"=as.factor(signif.OPEN))

```

Then coefficient plots are created with ggplot2

```{r Plotting source, include=FALSE}

#Plotting coefficient plot of source

df.source<-df.params[,c(2,6:7,14)]

level_order<-rownames(df.source[order(df.source$coef.source.GAL, decreasing = FALSE),])

colnames(df.source)<-c("Coefficient","Conf.low","Conf.high","Significance")


ggcoefplot.source<-ggplot(df.source, aes(x=Coefficient, y=factor(rownames(df.source),level=level_order),col=Significance)) +
  geom_vline(aes(xintercept=0), lty=2, alpha=0.5) +
  coord_cartesian(xlim=c(-2,2)) +
  geom_point() +
  scale_color_manual(values=c("darkseagreen4","dodgerblue3","grey")) +
  geom_segment(aes(x=Conf.low, xend=Conf.high, yend=factor(rownames(df.source)),colour=Significance)) +
  xlab("Coefficients") +
  ylab("Orders") +
  theme_bw()


ggcoefplot.source<-ggcoefplot.source +  theme(legend.position="top", legend.title = element_blank(), legend.direction ="vertical" )

ggcoefplot.source<-ggcoefplot.source +  theme(text = element_text(size = 14))
  
ggcoefplot.source


  
```


```{r Plotting season, include=FALSE}


#Plotting coefficient plot of season

df.season<-df.params[,which(grepl("season", colnames(df.params)))]

level_order1<-rownames(df.season[order(df.season$coef.season.MID, decreasing = FALSE),])
level_order2<-rownames(df.season[order(df.season$coef.season.LATE, decreasing = FALSE),])


colnames(df.season)<-c("Coefficient.MID","Coefficient.LATE","Conf.low.MID","Conf.high.MID","Conf.low.LATE","Conf.high.LATE","Significance", "Significance.LATE" )


ggcoefplot.season<-ggplot(df.season, aes(x=Coefficient.MID, y=factor(rownames(df.source),level=level_order1),col=Significance)) +
  geom_vline(aes(xintercept=0), lty=2, alpha=0.5) +
  coord_cartesian(xlim=c(-3,20)) +
  geom_point(position = position_nudge(x=0, y=0.1)) +
  geom_segment(aes(x=Conf.low.MID, xend=Conf.high.MID, yend=factor(rownames(df.source)),colour=Significance), position = position_nudge(x=0, y=0.1)) +
  
  scale_color_manual(values=c("#4fe7c7","#83370d","#fb7264","grey")) +
  geom_point(data=df.season, aes(x=Coefficient.LATE, y=factor(rownames(df.source),level=level_order),col=Significance.LATE), position = position_nudge(x=0, y=-0.1)) +
  geom_segment(data=df.season,aes(x=Conf.low.LATE, xend=Conf.high.LATE, yend=factor(rownames(df.source)),colour=Significance.LATE), position = position_nudge(x=0, y=-0.1)) +
  
  
  xlab("Coefficients") +
  ylab("Orders") +
  theme(legend.position="top") +
  theme_bw()

ggcoefplot.season<-ggcoefplot.season +  theme(legend.position="top", legend.title = element_blank())

ggcoefplot.season<-ggcoefplot.season +  theme(text = element_text(size = 14))
  
ggcoefplot.season

```



```{r Plotting habitat, include=FALSE}

#Plotting coefficient plot of habitat

df.habitat<-df.params[,which(grepl("OPEN", colnames(df.params)))]

level_order<-rownames(df.habitat[order(df.habitat$coef.habitat.OPEN, decreasing = FALSE),])

colnames(df.habitat)<-c("Coefficient","Conf.low","Conf.high","Significance")


ggcoefplot.habitat<-ggplot(df.habitat, aes(x=Coefficient, y=factor(rownames(df.habitat),level=level_order),col=Significance)) +
  geom_vline(aes(xintercept=0), lty=2, alpha=0.5) +
  coord_cartesian(xlim=c(-2,2)) +
  geom_point() +
  scale_color_manual(values=c("#AED0A5","#F9F0BC","grey")) +
  geom_segment(aes(x=Conf.low, xend=Conf.high, yend=factor(rownames(df.habitat)),colour=Significance)) +
  xlab("Coefficients") +
  ylab("Orders") +
  theme(legend.position="top") +
  theme_bw()

ggcoefplot.habitat<-ggcoefplot.habitat +  theme(legend.position="top", legend.title = element_blank(), legend.direction ="vertical" )

ggcoefplot.habitat<-ggcoefplot.habitat +  theme(text = element_text(size = 14))
  
ggcoefplot.habitat


  
```

# Arrange and export plots

```{r Export plots, echo=FALSE}

gg.tmp<-ggarrange(ggcoefplot.source,ggcoefplot.habitat, labels = c("A","B"), nrow = 1, ncol=2)

ggarrange(gg.tmp,ggcoefplot.season, labels = c("","C"), nrow = 2, ncol=1)
ggsave("COI/Local_analysis/output/plots/Coefficient_plots_gllvm.png", width=12, height=9)


```

```{r Export AIC values, echo=FALSE}

AICs<-as.data.frame(cbind(c("final model", "with.interactions","without.row.effects"),c(AICc(fitBEST), AICc(fitBEST.int), AICc(fitBEST.norow))))

colnames(AICs)<-c("model", "AIC")

write_delim(AICs, file="COI/Local_analysis/output/textfiles/AICs_gllvm.txt")

```

